cardinalR: Generating Interesting High-Dimensional Data Structures

### 2025-12-19 extra/paper-cardinalR-code.Rmd
Updated with new package versions

### Cone

To simulate a cone-shaped structure in arbitrary dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cone)"} else { "Figure \\@ref(fig:cone-proj)"})`), we define a function `gen_cone(n, p, h, ratio)`, which creates a high-dimensional cone with options for a sharp or blunted apex, allowing for a dense concentration of points near the tip.

This function generates $n$ points in $p\text{-}D$, where the last dimension, $X_p$, represents the height along the cone's axis, and the first $p-1$ dimensions define a shrinking hyperspherical cross-section toward the tip. Heights are sampled from a truncated exponential distribution, $X_p \sim \text{Exp}(\lambda = 2/h)$, truncated to the interval $[0, h]$, producing a higher density of points near the tip. At each height $X_p$, the radius of the cross-section increases linearly from base to tip according to $r = r_{\text{min}} + (r_{\text{max}} - r_{\text{min}}) X_p / h$, where $r_{\text{min}} = \text{ratio} \in [0, 1]$ and $r_{\text{max}} = 1$.

For each point, a direction in the first $p-1$ dimensions is sampled uniformly on a $(p-1)$-dimensional hypersphere using generalized spherical coordinates. The radial coordinates are scaled by the height-dependent radius $r$, producing the conical taper. In three dimensions ($p = 3$), this results in a classical $3\text{-}D$ cone, while for $p > 3$, additional dimensions provide a smooth embedding into higher-dimensional space, preserving the conical structure.

Cone-shaped structures appear in particle dispersions, light beams, and tapering processes, where spread decreases along one axis. They are also used to benchmark clustering and dimensionality reduction methods [@hadsell2006].

```{r cone-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from the $3\\text{-}D$ \\texttt{cone} data. Points are concentrated near the tip along the height dimension, while the radius of the hyperspherical cross-section decreases linearly toward the apex. These projections show how the conical geometry is preserved.", fig.pos="H", fig.pos="!ht", fig.width=12, fig.height=4}
cone <- gen_cone(n = 1000, p = 4, h = 5, ratio = 0.5)

@@

# Function to scale data manually
# scale_data_manual <- function(data) {
#   # Step 1: Center the data (mean 0)
#   data_centered <- center_data(data)
#
#   # Step 2: Calculate the standard deviation of each dimension
#   sds <- apply(data_centered, 2, sd)
#
#   # Step 3: Scale each dimension to have the range [0, 1]
#   data_scaled <- apply(data_centered, 2, function(col) col / max(abs(col)))
#
#   # Step 4: Scale dimensions according to their variation
#   # The dimension with the highest standard deviation is scaled to [-1, 1]
#   # Other dimensions are scaled to smaller ranges based on their standard deviations
#   max_sd <- max(sds)
#
#   # Normalize the standard deviations to get scaling factors
#   scaling_factors <- sds / max_sd
#
#   for (i in seq_along(scaling_factors)) {
#     data_scaled[, i] <- data_scaled[, i] * scaling_factors[i]
#   }
#
#   # Combine the scaled data with the 'type' column and return as a tibble
#   data_scaled <- as_tibble(data_scaled)
#
#   return(data_scaled)
# }

@@

centered_data <- center_data(cone)

## First projection
projection <- cbind(
  c(0.22417,-0.02943,-0.02870,0.08418),
  c(-0.02715,0.18019,0.05053,0.15254))

proj_obj1 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.6,
                                              axis_scaled = 3, 
                                              axis_pos_x = -0.4, 
                                              axis_pos_y = -0.4, 
                                              threshold = 0.053))

cone_proj1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.5, 0.7), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

## Second projection
projection <- cbind(
    c(0.15591,0.17907,0.03198,0.04496),
    c(-0.15376,0.10435,0.15766,0.00547))

proj_obj2 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.3,
                                              axis_scaled = 2, 
                                              axis_pos_x = -0.2, 
                                              axis_pos_y = -0.2, 
                                              threshold = 0.02))

cone_proj2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.25, 0.35), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

## Third projection
projection <- cbind(
    c(-0.13548,0.13832,0.05991,-0.13544),
    c(0.08479,-0.10719,-0.01572,-0.20123))

proj_obj3 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.6,
                                              axis_scaled = 3, 
                                              axis_pos_x = -0.6, 
                                              axis_pos_y = -0.6, 
                                              threshold = 0.04))

cone_proj3 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.7, 0.5), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

cone_proj1 + cone_proj2 + cone_proj3 +
  plot_layout(ncol = 3, guides = "collect") 
```

>>>>

### Branching

A branching structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology [@trapnell2014]. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, total number of points (`n`) generated across all branches, with points allocated approximately evenly among branches, and number of branches (`k`). Although these functions can generate multiple branches, they do not produce a formal *multicluster* dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters.

The simplest structures are approximately linear branches in $2\text{-}D$, generated by the `gen_linearbranches(n, k)` function. These consist of $k$ short line segments in the first two dimensions, with added jitter to simulate variability. Mathematically, each branch $i$ is defined as

$$
X_1 \sim U(a_i, b_i), \quad X_2 = s_i (X_1 - x_{\text{start},i}) + y_{\text{start},i} + \epsilon, \quad \epsilon \sim U(0, \delta),
$$

where $(x_{\text{start},i}, y_{\text{start},i})$ is the starting point of branch $i$, $\delta$ controls local jitter, and $s_i$ is the slope, initialized as

$$
s_i =
\begin{cases}
0.5 & i = 1, \\
-0.5 & i = 2, \\
\text{randomly sampled from } [s_{\min}, s_{\max}] & i = 3, \dots, k.
\end{cases}
$$

The jitter term is sampled from a one-sided uniform distribution to introduce directional variability without altering branch orientation.

Branches $1$ and $2$ are initialized with fixed slopes and starting points, while later branches are iteratively added at locations chosen to avoid overlap with existing branches, producing a set of connected linear paths.

To introduce curvature, the `gen_curvybranches(n, k)` function generates $k$ curvilinear branches in $2\text{-}D$. Each branch follows a quadratic trajectory of the form

$$
X_1 \sim U(a_i, b_i), \quad X_2 = 0.1 X_1 + s_i X_1^2 + \epsilon, \quad \epsilon \sim U(-\delta, \delta),
$$
where $(a_i, b_i)$ defines the domain of the branch, $s_i$ controls curvature, and $\delta$ introduces local jitter. For the first two branches, the parameters are fixed to establish reference shapes: $(a_1,b_1,s_1) = (0,1,1), \quad (a_2,b_2,s_2) = (-1,0,-2)$. Additional branches are attached iteratively to existing structures. Each new branch $i$ starts at a selected point $(x_{\text{start},i}, y_{\text{start},i})$ from the current structure and extends according to

$$
X_1 \sim U(x_{\text{start},i}, x_{\text{start},i}+1), \quad X_2 = 0.1 X_1 - s_i (X_1^2 - x_{\text{start},i}) + y_{\text{start},i},
$$

where $s_i$ is a scale factor controlling the curvature of branch $i$. For the first few initial branches, $s_i$ can be fixed (e.g., $s_1 = 1, s_2 = 2$) to establish reference shapes, while for subsequent branches it is sampled from a predefined set, such as $s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5\}$, to control curvature magnitude and direction.

The `gen_expbranches(n, k)` function creates $k$ exponential branches in $2\text{-}D$, radiating from a central region. Each branch $i$ is defined as

$$
X_1 \sim U(-2,2), \quad X_2 = \exp(\sigma_i \, s_i \, X_1) + \epsilon, \quad \epsilon \sim U(0, \delta), \quad s_i \sim U(0.5,2),
$$

where $\sigma_i = (-1)^{i+1}$ alternates the sign of the exponent to produce mirror-symmetric branches. The parameter $s_i$ controls the steepness of branch $i$, and $\delta$ introduces small local jitter.

High-dimensional generalizations are provided by `gen_orglinearbranches(n, p, k)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) and `gen_orgcurvybranches(n, p, k)`. For branch 
$i$, the active coordinate pair $(i_1, i_2)$ indexes the selected $2\text{-}D$ subspace. When `allow_share = TRUE`, multiple branches may share the same subspace; otherwise, subspaces are sampled without replacement until all possible $\binom{p}{2}$ combinations are exhausted, after which additional branches may repeat subspaces.

In both cases, branch $i$ is generated according to

$$
X_{i_1} \sim U(a_i, b_i), \quad
X_{i_2} = f_i(X_{i_1}) + \epsilon, \quad
\epsilon \sim N(0, \sigma^2),
$$
where $a_i$ and $b_i$ define the domain of the branch and $\epsilon$ introduces smooth variability in the $p\text{-}D$ space. The function $f_i(\cdot)$ determines the branch geometry:
$$
f_i(x) =
\begin{cases}
s_i x, & \text{linear branches}, \\
-s_i x^2, & \text{curvilinear branches}.
\end{cases}
$$

The scale factor $s_i$ controls slope (linear branches) or curvature (curvilinear branches) and is assigned as follows: for the first $\binom{p}{2}$ branches, $s_i = 1$; for additional branches when $k > \binom{p}{2}$, $s_i$ is randomly drawn from the set $\{1, 1.5, 2, \dots, 8\}$.

Across all branching generators, the scale parameter $s_i$ controls the strength of deviation from linearity, determining slope, curvature, or growth rate depending on the branch geometry.

```{r branch-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from the $4\\text{-}D$ \\texttt{orgcurvybranches} data. Each shows a different projection, illustrating how the linear branches appear from multiple viewing angles. These views highlight the dataset’s underlying branching structure and demonstrate how projections reveal patterns that are otherwise hidden in higher dimensions.", fig.width=12, fig.height=4, fig.pos="H", fig.pos="!ht"}

set.seed(20240412)

@@

@@

orglinearbranches <- gen_orglinearbranches(n = 1000, p = 4, k = 4)

## First projection
projection <- cbind(
    c(-0.083573,-0.019043,0.006574,-0.035774),
    c(0.005380,0.068541,-0.030810,-0.054717))

proj_obj1 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.5,
                                              axis_scaled = 10, 
                                              axis_pos_x = -0.5, 
                                              axis_pos_y = -0.5, 
                                              threshold = 0.05))

orglinearbranches_proj1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.6, 0.4), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 4,
  is_color = FALSE)

## Second projection
projection <- cbind(
  c(-0.06614,0.15134,-0.44278,0.27367),
  c(-0.50493,0.14388,0.14615,0.03486))

proj_obj2 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.9,
                                              axis_scaled = 1, 
                                              axis_pos_x = -0.85, 
                                              axis_pos_y = -0.85, 
                                              threshold = 0.05))

orglinearbranches_proj2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-1, 1), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 4,
  is_color = FALSE)

## Third projection
projection <- cbind(
  c(0.10704,0.04822,-0.03416,0.53224),
  c(0.50329,0.07199,-0.16064,-0.11806))

proj_obj3 <- get_projection(projection = projection, 
                            centered_data = centered_data, 
                            axis_param = list(limits = 0.9,
                                              axis_scaled = 1.5, 
                                              axis_pos_x = -0.85, 
                                              axis_pos_y = -0.85, 
                                              threshold = 0.1))

orglinearbranches_proj3 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-1, 1), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 4,
  is_color = FALSE)

orglinearbranches_proj1 + orglinearbranches_proj2 + orglinearbranches_proj3 +
  plot_layout(ncol = 3, guides = "collect") 
```

### 2025-12-19 paper-cardinalR.tex
Updated with new package versions

{\centering \includegraphics[width=0.8\linewidth,alt={Workflow for generating high-dimensional clustered data, from specifying input parameters to producing the final labeled dataset with optional noise dimensions.}]{figures/cardinalR_workflow}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={The figure shows a 2-D scatter plot of projected branch structures in a high-dimensional space. The horizontal axis represents one coordinate (first index of the active pair) and the vertical axis represents the other coordinate (second index of the active pair), each spanning a moderate, symmetric range around zero (for example, roughly −2 to 2). Points are grouped visually into several distinct linear or gently curving branches, each corresponding to a different active 2-D subspace; these branches appear as separate streaks or arcs radiating through the plane. Colour (or another visual attribute such as line/point grouping) is used to distinguish branches that occupy different subspaces, and some branches may overlap or run close together where subspaces are shared. The main pattern is a set of well-separated, structured branches with little noise, illustrating how high-dimensional linear or curvy branches project into 2-D coordinate pairs.}]{paper-cardinalR_files/figure-latex/branch-proj-1} \caption{Three $2\text{-}D$ projections from the $4\text{-}D$ \texttt{orgcurvybranches} data. Each shows a different projection, illustrating how the linear branches appear from multiple viewing angles. These views highlight the dataset’s underlying branching structure and demonstrate how projections reveal patterns that are otherwise hidden in higher dimensions.}\label{fig:branch-proj}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={ A 2D scatterplot shows the projection of a high-dimensional cone-shaped point cloud. The horizontal and vertical axes are unlabeled numeric coordinates, each spanning a roughly symmetric range around zero (for example, from about -1 to 1). Hundreds of points form a dense cluster near one end of the plot, corresponding to the cone’s apex, with density gradually decreasing as points extend outward in a flared, cone-like or triangular region. There are no colours, shapes, or facets used; all points share the same appearance, and the key visible pattern is the concentrated tip transitioning smoothly into a wider spread of points, illustrating a cone structure in projection.}]{paper-cardinalR_files/figure-latex/cone-proj-1} \caption{Three $2\text{-}D$ projections from the $3\text{-}D$ \texttt{cone} data. Points are concentrated near the tip along the height dimension, while the radius of the hyperspherical cross-section decreases linearly toward the apex. These projections show how the conical geometry is preserved.}\label{fig:cone-proj}
\end{figure}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Three $2\text{-}D$ projections from each of four $4\text{-}D$ pyramid-like datasets: pyrrect, pyrtri, pyrstar, and pyrholes, showing their distinct geometries, branching, and hollow regions.}]{paper-cardinalR_files/figure-latex/pyr-proj-1}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Three $2\text{-}D$ projections from each of three datasets: circle, curvycycle, and clusteredspheres, showing closed loops, twisted rings, and compact spherical clusters, illustrating projection effects on structure, density, and separation.}]{paper-cardinalR_files/figure-latex/sphere-proj-1}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Three $2\text{-}D$ projections from each of two trefoil knot datasets, showing complex $4\text{-}D$ trefoil loops and simpler 3-D knot structures, illustrating curvature, twisting, and separation across dimensions.}]{paper-cardinalR_files/figure-latex/trefoil-proj-1}

@@

{\centering \includegraphics[width=0.7\linewidth,alt={Three $2\text{-}D$ projections from each of four $4\text{-}D$ datasets: curvycylinder, sphericalspiral, conicspiral, and nonlinear, showing cylindrical twists, spiral and conical helices, and warped $2\text{-}D$ surfaces, illustrating curvature, density, and continuity.}]{paper-cardinalR_files/figure-latex/triginometric-proj-1}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={Three $2\text{-}D$ projections of five $4\text{-}D$ synthetic clusters: helical spiral (dark green), hemisphere (orange), uniform cube (purple), cone (pink), and Gaussian (light green), showing spatial separation and cluster geometry.}]{paper-cardinalR_files/figure-latex/highd-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the five clusters data. The helical spiral cluster is represented in dark green, the hemisphere cluster in orange, the uniform cube-shaped cluster in purple, the blunted cone cluster in pink, and the Gaussian-shaped cluster in light green.}\label{fig:highd-proj}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={Composite plot assessing six NLDR layouts of five clusters data using HBE over varying binwidths, showing tSNE, UMAP, PAHTE, TriMAP, PaCMAP, and PCA. Layouts a and b are optimal, f is poor, with color-coded lines matching the scatterplots of each layout.}]{paper-cardinalR_files/figure-latex/fig-nldr-layouts-1} \caption{Assessing which of the 6 NLDR layouts ((a) tSNE, (b) UMAP, (c) PAHTE, (d) TriMAP, (e) PaCMAP, and (f) PCA) of the five clusters data is the better representation using HBE for varying binwidth ($a_1$). Colour is used for the lines and points in the left plot to match the scatterplots of the NLDR layouts (a-f). Layout f is universally poor. Layouts a and b are universally optimal. Layout b shows six well-separated clusters and layout a shows close clusters, thus layout a is the best choice.}\label{fig:fig-nldr-layouts}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cluster-stats-1} \caption{Cluster validity metrics for solutions with $2-10$ clusters obtained using $k$-means, hierarchical, and model-based clustering. Several indices consistently suggest that $4-5$ clusters provide the best balance of separation and compactness, with $k$-means performing slightly better across metrics.}\label{fig:fig-cluster-stats}
\end{figure}

Figure \ref{fig:fig-cluster-stats} shows a selection of cluster metrics for \(2-10\) clusters for each of the methods, \(k\)-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a \(4-5\) cluster solution. Pearson gamma (\texttt{pearsongamma}) increases sharply up to five clusters before levelling off, Calinski--Harabasz index (\texttt{ch}) increases sharply from 4 to 5 clusters and Dunn (\texttt{dunn2}) has a maximum at 5 for two methods and at 4 for \(k\)-means. All of these are interpreted as higher is better. With the other three, lower is better. WB ratio (\texttt{wb.ratio}) and within-cluster sum of squares (\texttt{within.cluster.ss}) steadily decline with number of clusters, possibly elbowing around 5 clusters. The S-index (\texttt{sindex}) is optimised at 4 clusters for \(k\)-means, 3, 6 or 8 for hierarchical clustering, and 4 or 8 for model-based. Overall, \(k\)-means performs slightly better than the hierarchical and model-based clustering across most metrics and number of clusters.

\begin{figure}[!ht]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/highd-proj-clust-algo-pdf-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the five clusters data coloured by the $k$-means four- (a1-a3) and five-cluster (b1-b3) solutions. The intermixing of colours within each projection reflects misclassification in both solutions, showing the difficulty of using $k$-means to capture the dataset’s nonlinear and heterogeneous shapes.}\label{fig:highd-proj-clust-algo-pdf}
\end{figure}

>>>>

{\centering \includegraphics[width=0.8\linewidth,alt={The figure is a workflow diagram showing the process for generating high-dimensional clustered data. Boxes or nodes represent steps such as specifying input parameters (number of points, number of clusters, cluster locations, shapes, scaling, rotations, and optional background noise), generating each cluster with a shape generator, optionally rotating or scaling clusters, combining them into a single dataset, adding background noise if desired, and labeling each observation by cluster shape. Arrows indicate the flow of the process from parameter specification to the final labeled dataset. The diagram uses a clear layout with distinct steps, showing how data are constructed in sequence.}]{figures/cardinalR_workflow}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={The figure consists of three 2‑D scatter plots, each showing a different 2‑D projection from the 4‑D orgcurvybranches dataset. In each plot, the horizontal axis represents one coordinate of the selected 2‑D projection, and the vertical axis represents the other coordinate. Points are plotted as individual markers and appear grouped into multiple elongated clusters or branches, which curve or radiate through the plot. Each branch is distinguished by colour, with overlapping branches occurring when subspaces are shared. The plots have roughly square aspect ratios, and together they show how the same high‑dimensional branching structure appears differently from three viewing angles.}]{paper-cardinalR_files/figure-latex/branch-proj-1} \caption{Three $2\text{-}D$ projections from the $4\text{-}D$ \texttt{orgcurvybranches} data. Each shows a different projection, illustrating how the linear branches appear from multiple viewing angles. These views highlight the dataset’s underlying branching structure and demonstrate how projections reveal patterns that are otherwise hidden in higher dimensions.}\label{fig:branch-proj}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={The figure consists of three 2‑D scatter plots, each showing a different 2‑D projection of the 4‑D cone dataset. In each plot, the horizontal and vertical axes represent two continuous numeric dimensions. Points are plotted individually and are concentrated near one end, forming a narrow tip, and gradually spread out along the other direction, creating a wider base. The points trace an elongated, triangular or cone-like shape in each projection. All points use the same colour and marker, with no additional visual encodings, and the plots have roughly square aspect ratios. Together, the three projections illustrate the distribution of points along the conical structure from different viewing angles.}]{paper-cardinalR_files/figure-latex/cone-proj-1} \caption{Three $2\text{-}D$ projections from the $4\text{-}D$ \texttt{cone} data. Points are concentrated near the tip along the height dimension, while the radius of the hyperspherical cross-section decreases linearly toward the apex. These projections show how the conical geometry is preserved.}\label{fig:cone-proj}
\end{figure}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={The figure shows three 2‑D scatter plots for each of four 4‑D datasets: pyrrrect, pyrtri, pyrstar, and pyrholes. In each plot, the horizontal and vertical axes represent continuous numeric coordinates of the selected 2‑D projection, and points are plotted individually as small, uniformly coloured markers with no additional visual encodings. In pyrrrect, points form a dense rectangular base tapering to a narrow tip; in pyrtri, points form a triangular distribution with sharper edges; in pyrstar, points extend into multiple pointed branches radiating from a central region; and in pyrholes, points cluster in a compact shape but leave hollow or open regions within it. The plots maintain roughly square aspect ratios and illustrate how these different pyramid-like structures appear when projected from four to two dimensions.}]{paper-cardinalR_files/figure-latex/pyr-proj-1}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={A 3D scatterplot shows a projection of points sampled from a 4‑dimensional hemisphere onto three coordinate axes. Each point represents a location on the surface of a 4‑D unit sphere restricted to one hemisphere. The horizontal (x) and vertical (y) axes, along with depth (z), span roughly from –1 to 1, reflecting the Cartesian coordinates of the projected points. The dots are uniformly scattered over a curved, bowl-like region rather than filling a full sphere, with no strong clusters or gaps, illustrating that the sampling covers only one side of the sphere. If colour is used, it likely encodes an additional coordinate or density but does not introduce distinct categorical groups.}]{paper-cardinalR_files/figure-latex/sphere-proj-1}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={The figure shows multiple scatterplots displaying two-dimensional projections of trefoil-shaped data derived from four- and three-dimensional spaces. In panels a1–a3, the trefoil4d data are shown as three different 2-D projections, where the horizontal and vertical axes represent linear combinations of the four original variables. The projected points form smooth, continuous looping curves that intersect and overlap in places, creating knot-like shapes whose apparent crossings and separations vary across projections. In panels b1–b3, the trefoil3d data are shown in three 2-D projections. The points trace a compact, closed loop with a characteristic trefoil-like knot structure, appearing smoother and less spread out than in the four-dimensional case. The figure highlights how the same underlying knot structure appears differently under changes in projection and dimensionality.}]{paper-cardinalR_files/figure-latex/trefoil-proj-1}

@@

{\centering \includegraphics[width=0.7\linewidth,alt={The figure consists of four rows of panels, each displaying three separate 2-D projection scatterplots from a four-dimensional synthetic dataset. Panels (a1–a3) show the curvycylinder dataset, where points form elongated, band-like structures that bend smoothly across projections, appearing cylindrical with gentle nonlinear curvature. Panels (b1–b3) show the sphericalspiral dataset, with points tracing curved, looping paths that resemble spiral arcs on a rounded surface, changing orientation across projections. Panels (c1–c3) show the conicspiral dataset, where point clouds expand outward in some projections while twisting along curved trajectories, creating tapered, spiral-like shapes. Panels (d1–d3) show the nonlinear dataset, in which points form warped surfaces with alternating tight and spread regions, visible as oscillating bands and smooth waves depending on the projection. Across all datasets, point density, curvature, and apparent continuity vary between projections, illustrating how different geometric structures appear under multiple 2-D views of the same four-dimensional data.}]{paper-cardinalR_files/figure-latex/triginometric-proj-1}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={The figure shows three 2-D projection plots of synthetic 4-D data containing five clusters with distinct geometric structures. The horizontal and vertical axes represent two projected dimensions, and no explicit units are shown. Points are coloured by cluster, with distinct colours representing the helical spiral, hemisphere, uniform cube, cone, and Gaussian clusters. The helical spiral cluster appears as an elongated, twisted band; the hemisphere cluster forms a curved, arc-like concentration; the uniform cube cluster fills a roughly rectangular region with fairly even density; the cone cluster exhibits a density gradient, narrowing at one end and spreading out at the other; and the Gaussian cluster forms a compact, approximately circular group near the centre of the plot. The clusters appear very close to one another in these projections, illustrating the difficulty of distinguishing cluster structure when projections bring clusters into close proximity.}]{paper-cardinalR_files/figure-latex/highd-proj-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the five clusters data. The helical spiral cluster is represented in dark green, the hemisphere cluster in orange, the uniform cube-shaped cluster in purple, the blunted cone cluster in pink, and the Gaussian-shaped cluster in light green.}\label{fig:highd-proj}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={A multi-panel figure compares 2-D scatterplots from six nonlinear dimensionality reduction methods applied to the same dataset with five true clusters. Each panel plots the first embedding dimension on the horizontal axis and the second embedding dimension on the vertical axis. Points represent individual observations and are coloured by cluster membership, with six distinct cluster colours reused consistently across panels. In the tSNE panel, clusters form compact, clearly separated groups with small gaps between clusters, indicating strong preservation of both local neighbourhoods and the global cluster layout. UMAP and PaCMAP also show six visibly distinct clusters that are moderately well separated but with slightly more overlap and distortion than tSNE. PHATE produces curved, nonlinear cluster shapes where clusters are stretched or intertwined, obscuring the original simple cluster geometry. TriMAP collapses the data into three main visible groups instead of six, with only small distances between these groups, suggesting loss of finer cluster structure. PCA displays the weakest structure: clusters overlap substantially and align along a roughly linear or planar trend, failing to reflect the underlying non-linear separation among the six groups.}]{paper-cardinalR_files/figure-latex/fig-nldr-layouts-1} \caption{Assessing which of the 6 NLDR layouts ((a) tSNE, (b) UMAP, (c) PAHTE, (d) TriMAP, (e) PaCMAP, and (f) PCA) of the five clusters data is the better representation using HBE for varying binwidth ($a_1$). Colour is used for the lines and points in the left plot to match the scatterplots of the NLDR layouts (a-f). Layout f is universally poor. Layouts a and b are universally optimal. Layout b shows six well-separated clusters and layout a shows close clusters, thus layout a is the best choice.}\label{fig:fig-nldr-layouts}
\end{figure}

@@

\begin{figure}[!]
\includegraphics[width=1\linewidth,alt={Multi-panel line chart showing six cluster quality metrics across numbers of clusters (x-axis: 2 to 10 clusters) for three clustering methods (k-means, hierarchical, model-based). Each panel displays one metric with the y-axis scaled to that metric’s values (not explicitly labeled in the text, but each metric is treated as higher-is-better or lower-is-better as described). Within each panel, three lines (one per method) trace the metric value as the number of clusters increases. For Pearson gamma, values for all three methods rise steeply up to about 5 clusters and then level off. For the Calinski–Harabasz index, values increase sharply from 4 to 5 clusters. For Dunn, the k-means line peaks around 4 clusters, while the hierarchical and model-based lines peak around 5. For WB ratio and within-cluster sum of squares, all three methods show a generally monotonic decline as the number of clusters increases, with a visible bend or elbow around 5 clusters. For the S-index, the k-means line reaches its best value around 4 clusters, the hierarchical line around 3, 6, or 8 clusters, and the model-based line around 4 or 8. Across panels, the k-means line is typically at or near the most favorable values for each metric, and several metrics simultaneously favor solutions with about 4–5 clusters.}]{paper-cardinalR_files/figure-latex/fig-cluster-stats-1} \caption{Cluster validity metrics for solutions with $2-10$ clusters obtained using $k$-means, hierarchical, and model-based clustering. Several indices consistently suggest that $4-5$ clusters provide the best balance of separation and compactness, with $k$-means performing slightly better across metrics.}\label{fig:fig-cluster-stats}
\end{figure}

Figure \ref{fig:fig-cluster-stats} shows a selection of cluster metrics for \(2-10\) clusters for each of the methods, \(k\)-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a \(4-5\) cluster solution. Pearson gamma (\texttt{pearsongamma}) increases sharply up to five clusters before leveling off, Calinski--Harabasz index (\texttt{ch}) increases sharply from 4 to 5 clusters and Dunn (\texttt{dunn2}) has a maximum at 5 for two methods and at 4 for \(k\)-means. All of these are interpreted as higher is better. With the other three, lower is better. WB ratio (\texttt{wb.ratio}) and within-cluster sum of squares (\texttt{within.cluster.ss}) steadily decline with number of clusters, possibly elbowing around 5 clusters. The S-index (\texttt{sindex}) is optimized at 4 clusters for \(k\)-means, 3, 6 or 8 for hierarchical clustering, and 4 or 8 for model-based. Overall, \(k\)-means performs slightly better than the hierarchical and model-based clustering across most metrics and number of clusters.

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={The figure shows three row-wise 2-D projection plots of synthetic 4-D data for two different k-means clustering solutions: four clusters in one panel and five clusters in the other. In each panel, the horizontal and vertical axes represent linear projections of the original high-dimensional variables, and each point corresponds to a data observation. Points are coloured by assigned cluster ID (four colours in the top row and five in the bottom row). In both solutions, clusters appear as roughly compact groups, but none aligns with the underlying geometric shapes in the data: each of the five true shapes is split across multiple clusters, and no cluster cleanly isolates a single shape. The four-cluster solution merges some shapes more strongly, while the five-cluster solution yields a finer partition that still fragments each shape; additional clusters beyond five would be needed to better match the original geometric structure.}]{paper-cardinalR_files/figure-latex/highd-proj-clust-algo-pdf-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the five clusters data coloured by the $k$-means four- (a1-a3) and five-cluster (b1-b3) solutions. The intermixing of colours within each projection reflects misclassification in both solutions, showing the difficulty of using $k$-means to capture the dataset’s nonlinear and heterogeneous shapes.}\label{fig:highd-proj-clust-algo-pdf}
\end{figure}
